```{r} 
library(tidyverse) 
```

``` {r}
mef_variants_supp <- read_tsv('D:/BirkBeck/Project/Ultralong/20_per_read_variant_calls.txt',
                     col_types=cols()) %>%
  filter(chrm=="KY962518.1_looped_2120_20_units") %>%
  select(-chrm) %>%
  mutate(pos=pos+1,
         unit_num=as.integer( ( pos - 1) / 44838) + 1,
         pos_ref=adjust_pos((pos %% 44838), 2120)) %>%
  group_by(read_id, strand, ref_log_prob, alt_log_prob,
           ref_seq, alt_seq, var_id, pos_ref) %>%
  summarise(unit_num=min(unit_num), .groups='drop') %>%
  select(read_id, pos_ref, ref=ref_seq, ref_prob=ref_log_prob,
         alt=alt_seq, alt_prob=alt_log_prob, unit_num) %>%
  pivot_longer(c(ref_prob, alt_prob), values_to='prob', names_to='dominant') %>%
  filter(prob>log(0.9)) %>%
  group_by(read_id, pos_ref, ref, alt, unit_num) %>%
  filter(n()==1) %>%
  mutate(dominant=gsub('_prob', '', dominant),
         allele=if_else(dominant=="ref", ref, alt)) %>%
  ungroup() %>%
  select(read_id, pos_ref, allele, unit_num) %>%
  distinct() %>%
  group_by(read_id, unit_num) %>%
  filter(n_distinct(pos_ref)==n()) %>%
  group_by(pos_ref) %>%
  filter(n_distinct(allele) > 1) %>%
  mutate(allele_num=dense_rank(allele)) %>%
  ungroup()
```

``` {r}
mef_variants_supp %>% 
  group_by(read_id, pos_ref) %>%
  filter(n_distinct(unit_num) > 1) %>%
  group_by(read_id, pos_ref, allele_num) %>% 
  summarise(num_allele = n()) %>%
  group_by(read_id, pos_ref) %>%
  mutate(freq_allele = num_allele / sum(num_allele)) %>%
  summarise(min_freq_allele = min(freq_allele), max_freq_allele = max(freq_allele)) %>%
  arrange(min_freq_allele)

```
```{r}

mef_variants_supp_counts <- mef_variants_supp %>% 
  group_by(read_id) %>% 
  mutate(num_units=n_distinct(unit_num)) %>%
  filter(num_units > 1)

```

```{r}

mef_variants_supp_multi <- mef_variants_supp %>% 
  group_by(read_id, unit_num) %>% 
  filter(! all(pos_ref > 0), ! all(pos_ref < 7980)) %>% 
  group_by(read_id) %>% 
  mutate(num_units=n_distinct(unit_num)) %>% 
  filter(num_units > 1)

```

``` {r}
mef_variants_supp_multi %>% 
  group_by(read_id, pos_ref) %>%
  filter(n_distinct(unit_num) > 1) %>%
  group_by(read_id, pos_ref, allele_num) %>% 
  summarise(num_allele = n()) %>%
  group_by(read_id, pos_ref) %>%
  mutate(freq_allele = num_allele / sum(num_allele)) %>%
  summarise(min_freq_allele = min(freq_allele), max_freq_allele = max(freq_allele)) %>%
  arrange(min_freq_allele)

```

#average methlyation visualisation
``` {r}
adjust_pos <- function(pos, loop_size=2120) {
  if_else(pos<=loop_size, pos-loop_size-1, pos-loop_size)
}

meth_calls <- read_tsv(
  'D:/Birkbeck/Project/Ultralong/20_per_read_modified_base_calls.txt', col_types=cols()) %>%
  filter(chrm=="KY962518.1_looped_2120_20_units") %>%
  select(-chrm) %>%
  mutate(pos=if_else(strand=='+', pos+1, pos),
         unit_num=as.integer( ( pos - 1) / 44838) + 1,
         meth_ref=adjust_pos((pos %% 44838), 2120)) %>%
  group_by(read_id, strand, can_log_prob, mod_log_prob, meth_ref) %>%
  summarise(unit_num=min(unit_num), .groups='drop') %>%
  select(read_id, meth_ref, can_prob=can_log_prob,
         mod_prob=mod_log_prob, unit_num) %>%
  pivot_longer(c(can_prob, mod_prob), values_to='prob', names_to='dominant') %>%
  filter(prob>log(0.9)) %>%
  group_by(read_id, meth_ref, unit_num) %>%
  distinct() %>%
  filter(n()==1) %>%
  mutate(meth=if_else(dominant=="mod_prob", 1, 0)) 

  multi_meth_calls <- meth_calls %>%
  group_by(read_id, unit_num) %>%
  filter(! all(meth_ref > 0), ! all(meth_ref < 7980)) %>%
  group_by(read_id) %>%
  filter(n_distinct(unit_num) > 1) %>%
  select(read_id, meth_ref, unit_num, meth)

meth_graph <- multi_meth_calls %>% 
  filter(meth_ref > -1000, meth_ref <= 13332) %>% 
  group_by(read_id, unit_num) %>% 
  summarise(mean_meth = mean(meth)) %>%
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, fill=mean_meth), shape=21, size=3) + 
  scale_fill_gradient2("Mean\nmethylation\n10 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5))
```

#allele visualisation
``` {r}
adjust_pos <- function(pos, loop_size=2120) {
  if_else(pos<=loop_size, pos-loop_size-1, pos-loop_size)
}

allele_calls <- read_tsv('D:/BirkBeck/Project/Ultralong/20_per_read_variant_calls.txt',
                     col_types=cols()) %>%
  filter(chrm=="KY962518.1_looped_2120_20_units") %>%
  select(-chrm) %>%
  mutate(pos=pos+1,
         unit_num=as.integer( ( pos - 1) / 44838) + 1,
         pos_ref=adjust_pos((pos %% 44838), 2120)) %>%
  group_by(read_id, strand, ref_log_prob, alt_log_prob,
           ref_seq, alt_seq, var_id, pos_ref) %>%
  summarise(unit_num=min(unit_num), .groups='drop') %>%
  select(read_id, pos_ref, ref=ref_seq, ref_prob=ref_log_prob,
         alt=alt_seq, alt_prob=alt_log_prob, unit_num) %>%
  pivot_longer(c(ref_prob, alt_prob), values_to='prob', names_to='dominant') %>%
  filter(prob>log(0.9)) %>%
  group_by(read_id, pos_ref, ref, alt, unit_num) %>%
  distinct() %>%
  filter(n()==1) %>%
  mutate(dominant=gsub('_prob', '', dominant),
         allele=if_else(dominant=="ref", ref, alt)) %>%
  ungroup() %>%
  select(read_id, pos_ref, allele, unit_num) %>%
  distinct() %>%
  group_by(read_id, unit_num) %>%
  filter(n_distinct(pos_ref)==n()) %>%
  group_by(pos_ref) %>%
  filter(n_distinct(allele) > 1) %>%
  mutate(allele_num=dense_rank(allele)) %>%
  ungroup()

  multi_allele_calls <- allele_calls %>%
  group_by(read_id, unit_num) %>%
  filter(! all(pos_ref > 0), ! all(pos_ref < 7980)) %>%
  group_by(read_id) %>%
  filter(n_distinct(unit_num) > 1) %>%
  select(read_id, pos_ref, unit_num, allele)

allele_graph <- multi_allele_calls %>% 
  filter(pos_ref == 7980) %>% 
  group_by(read_id, unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele), size=3) + 
  scale_shape_manual(values=c('A'=65, 'C'=67, 'G'=71, 'T'=84)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5))
```
#mef_variants_supp %>% group_by(pos_ref, allele) %>% summarise(num_allele=n()) %>% group_by(pos_ref) %>% summarise(maf=min(num_allele)/sum(num_allele)) %>% filter(maf >0.4 & maf <0.6)
#to find positions with allele freuqency around 50%
```{r}
joint_calls <- meth_calls %>%
  group_by(read_id, unit_num) %>%
  summarise(mean_meth=mean(meth)) %>%
  inner_join(allele_calls, by=c('read_id','unit_num')) %>%
  group_by(read_id, unit_num) %>%
  filter(! all(pos_ref > 0), ! all(pos_ref < 7980)) %>%
  group_by(read_id) %>%
  filter(n_distinct(unit_num) > 1)

graph_7980 <- joint_calls %>% 
  filter(pos_ref == 7980) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number',y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) +
  ggtitle("Position 7980")

graph_10377 <- joint_calls %>% 
  filter(pos_ref == 10377) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) +
  ggtitle("Position 10377")

graph_10086 <- joint_calls %>% 
  filter(pos_ref == 10086) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) +
  ggtitle("Position 10086")

graph_12807 <- joint_calls %>% 
  filter(pos_ref == 12807) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) +
  ggtitle("Position 12807")

graph_10104 <- joint_calls %>% 
  filter(pos_ref == 10104) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) +
  ggtitle("Position 10104")

graph_11253 <- joint_calls %>% 
  filter(pos_ref == 11253) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) + 
  ggtitle("Position 11253")

graph_2298 <- joint_calls %>% 
  filter(pos_ref == 2298) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) + 
  ggtitle("Position 2298")

graph_5734 <- joint_calls %>% 
  filter(pos_ref == 5734) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) + 
  ggtitle("Position 5734")

graph_8782 <- joint_calls %>% 
  filter(pos_ref == 8782) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) + 
  ggtitle("Position 8782")

graph_12986 <- joint_calls %>% 
  filter(pos_ref == 12986) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) + 
  ggtitle("Position 12986")

graph_8069 <- joint_calls %>% 
  filter(pos_ref == 8069) %>% 
  group_by(read_id , unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele, fill=mean_meth), size=4) + 
  scale_shape_manual(values=c('A'=21, 'C'=23, 'G'=22, 'T'=24)) +
  scale_fill_gradient2("Mean\nmethylation\n20 Units", low="red", high="blue", mid="white", midpoint=0.5, limits=c(0,1)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5)) + 
  ggtitle("Position 8069")
```


#check how many real units in read_id fdcda3ed-4bc2-44e9-943c-a89de260674d
``` {r}
adjust_pos <- function(pos, loop_size=2120) {
  if_else(pos<=loop_size, pos-loop_size-1, pos-loop_size)
}

allele_calls <- read_tsv('D:/BirkBeck/Project/Ultralong/20_per_read_variant_calls.txt',
                     col_types=cols()) %>%
  filter(chrm=="KY962518.1_looped_2120_20_units") %>%
  filter(read_id=="fc1efbc6-4ed7-4767-b9a0-370c74ea625e") %>%
  select(-chrm) %>%
  mutate(pos=pos+1,
         unit_num=as.integer( ( pos - 1) / 44838) + 1,
         pos_ref=adjust_pos((pos %% 44838), 2120)) %>%
  group_by(read_id, strand, ref_log_prob, alt_log_prob,
           ref_seq, alt_seq, var_id, pos_ref) %>%
  summarise(unit_num=min(unit_num), .groups='drop') %>%
  select(read_id, pos_ref, ref=ref_seq, ref_prob=ref_log_prob,
         alt=alt_seq, alt_prob=alt_log_prob, unit_num) %>%
  pivot_longer(c(ref_prob, alt_prob), values_to='prob', names_to='dominant') %>%
  filter(prob>log(0.9)) %>%
  group_by(read_id, pos_ref, ref, alt, unit_num) %>%
  distinct() %>%
  filter(n()==1) %>%
  mutate(dominant=gsub('_prob', '', dominant),
         allele=if_else(dominant=="ref", ref, alt)) %>%
  ungroup() %>%
  select(read_id, pos_ref, allele, unit_num) %>%
  distinct() %>%
  group_by(read_id, unit_num) %>%
  filter(n_distinct(pos_ref)==n()) %>%
  group_by(pos_ref) %>%
  filter(n_distinct(allele) > 1) %>%
  mutate(allele_num=dense_rank(allele)) %>%
  ungroup()

  multi_allele_calls <- allele_calls %>%
  group_by(read_id, unit_num) %>%
  filter(! all(pos_ref > 0), ! all(pos_ref < 7980)) %>%
  group_by(read_id) %>%
  filter(n_distinct(unit_num) > 1) %>%
  select(read_id, pos_ref, unit_num, allele)

allele_graph <- multi_allele_calls %>% 
  filter(pos_ref == 7980) %>% 
  group_by(read_id, unit_num) %>% 
  ggplot() + 
  geom_point(aes(x=as.factor(unit_num), y=read_id, shape=allele), size=3) + 
  scale_shape_manual(values=c('A'=65, 'C'=67, 'G'=71, 'T'=84)) +
  theme_bw() +
  labs(x='Unit Number', y='') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5))
```



```{r}
coding_regions_h <- tribble(
  ~region,~start,~end,
  '18S',3658,5526,
  '5.8S',6597,6753,
  '28S',7921,12971)
# ADJUST POSITIONS - done when creating meth_calls and allele_calls
# group_by(pos, allele) %>%
# summarise(num_reads=n()) %>%
# mutate(read_freq=num_reads/sum(num_reads))

coding_reads <- meth_calls %>%
  group_by(read_id, unit_num) %>%
  summarise(mean_meth=mean(meth)) %>%
  inner_join(allele_calls, by=c('read_id','unit_num')) %>%
  group_by(read_id, unit_num) %>%
  filter(! all(pos_ref > 0), ! all(pos_ref < 7980)) %>%
  group_by(read_id) %>%
  filter(n_distinct(unit_num) > 1) %>%
  filter((pos_ref >= coding_regions_h$start[1] & pos_ref <= coding_regions_h$end[1]) | 
           (pos_ref >= coding_regions_h$start[2] & pos_ref <= coding_regions_h$end[2]) |
           (pos_ref >= coding_regions_h$start[3] & pos_ref <= coding_regions_h$end[3])) %>%
  group_by(pos_ref, allele) %>%
  summarise(num_reads=n()) %>%
  mutate(read_freq=num_reads/sum(num_reads)) %>%
  ggplot() + 
  geom_col(aes(x=as.factor(pos_ref), y=read_freq, fill=allele)) + 
  #facet_grid(rows = vars(read_freq)) + 
  theme_bw() +
  labs(x='Position', y='Read Frequency') +
  theme(panel.border=element_rect(size=1.1),
        axis.text.y=element_text(size=7.5))
```


